Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-14 paper/preprint/paper.tex

\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{counterfactual training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules that are carefully hand-crafted by humans, modern
machine learning (ML) models are tasked with learning representations
directly from data, guided by narrow objectives such as predictive
accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and
Courville 2016}). Modern advances in computing have made it possible to
provide such models with ever-growing degrees of freedom to achieve that
task, which frequently allows them to outperform traditionally more
parsimonious models. Unfortunately, in doing so, models learn
increasingly complex and highly sensitive representations that humans
can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
(DL) revolution, Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}) showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AEs):
perturbed versions of data instances that yield vastly different model
predictions despite being ``imperceptible'' in that they are
semantically indifferent from their factual counterparts. Even though
some partially effective mitigation strategies have been proposed---most
notably \textbf{adversarial training}
(\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
Szegedy 2014})---truly robust deep learning remains unattainable even
for models that are considered ``shallow'' by today's standards
(\citeproc{ref-kolter2023keynote}{Kolter 2023}).

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}) who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) who consider
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage CEs during
training and get adversarial examples essentially for free.

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
In this context, domain constraints can be imposed by simply projecting
counterfactuals back to the specified domain, if the previous gradient
step resulted in updated feature values that were out-of-domain.
Mutability constraints can similarly be enforced by setting partial
derivatives to zero to ensure that features are only perturbed in the
allowed direction, if at all.

@@

Our key outcome of interest is how well do models perform with respect
to explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain. We rely on the distance-based metric used by Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}),

@@

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a set of multiple
counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
squared population maximum mean discrepancy
(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).

@@

Table~\ref{tbl-main} presents our main empirical findings. The top five
rows show the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. The following row shows the reduction
in implausibility as measured by Equation~\ref{eq-impl-div} and
aggregated across all test specifications of \emph{ECCo}. The final two
rows show the test accuracies for the model trained with CT and
conventionally trained models (``vanilla'').

@@

\caption{\label{tbl-main}Key plausibility and predictive performance
metrics for all datasets. The top five rows show the percentage
reduction in implausibility according to Equation~\ref{eq-impl-dist} for
varying degrees of the energy penalty used for \emph{ECCo} at test time.
The following row shows the reduction in implausibility as measured by
Equation~\ref{eq-impl-div} and aggregated across all test specifications
of \emph{ECCo}. The final two rows show the test accuracies for the
model trained with CT and conventionally trained models (``vanilla'').}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
Recognition (CVPR)}, 10041--51.
\url{https://doi.org/10.1109/CVPR42600.2020.01006}.

\bibitem[\citeproctext]{ref-altmeyer2023explaining}
Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
JuliaCon Conferences}, 1:130. 1.

\bibitem[\citeproctext]{ref-altmeyer2024faithful}
Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
Conference on Artificial Intelligence}, 38:10829--37. 10.

\bibitem[\citeproctext]{ref-augustin2020adversarial}
Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
{``Adversarial Robustness on in-and Out-Distribution Improves
Explainability.''} In \emph{European Conference on Computer Vision},
228--45. Springer.

\bibitem[\citeproctext]{ref-balashankar2023improving}
Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
Chi, Jilin Chen, and Alex Beutel. 2023. {``Improving Classifier
Robustness Through Active Generative Counterfactual Data
Augmentation.''} In \emph{Findings of the Association for Computational
Linguistics: EMNLP 2023}, 127--39.

\bibitem[\citeproctext]{ref-bell2024fairness}
Bell, Andrew, Joao Fonseca, Carlo Abrate, Francesco Bonchi, and Julia
Stoyanovich. 2024. {``Fairness in Algorithmic Recourse Through the Lens
of Substantive Equality of Opportunity.''}
\url{https://arxiv.org/abs/2401.16088}.

@@

\bibitem[\citeproctext]{ref-du2019implicit}
Du, Yilun, and Igor Mordatch. 2020. {``Implicit Generation and
Generalization in Energy-Based Models.''}
\url{https://arxiv.org/abs/1903.08689}.

\bibitem[\citeproctext]{ref-franceschi2017forward}
Franceschi, Luca, Michele Donini, Paolo Frasconi, and Massimiliano
Pontil. 2017. {``{Forward and Reverse Gradient-Based Hyperparameter
Optimization}.''} In \emph{Proceedings of the 34th International
Conference on Machine Learning}, edited by Doina Precup and Yee Whye
Teh, 70:1165--73. Proceedings of Machine Learning Research. PMLR.
\url{https://proceedings.mlr.press/v70/franceschi17a.html}.

\bibitem[\citeproctext]{ref-frankle2018lottery}
Frankle, Jonathan, and Michael Carbin. 2019. {``The Lottery Ticket
Hypothesis: Finding Sparse, Trainable Neural Networks.''} In
\emph{International Conference on Learning Representations}.
\url{https://openreview.net/forum?id=rJl-b3RcF7}.

\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
Freiesleben, Timo. 2022. {``The Intriguing Relation Between
Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
Machines} 32 (1): 77--109.

\bibitem[\citeproctext]{ref-goodfellow2014explaining}
Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
{``Explaining and Harnessing Adversarial Examples.''}
\url{https://arxiv.org/abs/1412.6572}.

@@

@@

\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
Guidotti, Riccardo. 2022. {``Counterfactual Explanations and How to Find
Them: Literature Review and Benchmarking.''} \emph{Data Mining and
Knowledge Discovery}, 1--55.

\bibitem[\citeproctext]{ref-guo2023counternet}
Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
End-to-End Training of Prediction Aware Counterfactual Explanations.''}
In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
Association for Computing Machinery.
\url{https://doi.org/10.1145/3580305.3599290}.

@@

\bibitem[\citeproctext]{ref-joshi2019realistic}
Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
Actionable Explanations in Black-Box Decision Making Systems.''}
\url{https://arxiv.org/abs/1907.09615}.

\bibitem[\citeproctext]{ref-kolter2023keynote}
Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)},
xvi--. Los Alamitos, CA, USA: IEEE Computer Society.
\url{https://doi.org/10.1109/SaTML54575.2023.00009}.

\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2017.
{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
Ensembles.''} \emph{Advances in Neural Information Processing Systems}
30.

@@

\bibitem[\citeproctext]{ref-luu2023counterfactual}
Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
Training for Improving Robustness of Pre-Trained Language Models.''} In
\emph{Proceedings of the 37th Pacific Asia Conference on Language,
Information and Computation}, 881--88.

@@

\bibitem[\citeproctext]{ref-ross2021learning}
Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
International Conference on Neural Information Processing Systems}. NIPS
'21. Red Hook, NY, USA: Curran Associates Inc.

\bibitem[\citeproctext]{ref-sauer2021counterfactual}
Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
Networks.''} \url{https://arxiv.org/abs/2101.06046}.

>>>>

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules that are carefully hand-crafted by humans, modern
machine learning (ML) models are tasked with learning representations
directly from data, guided by narrow objectives such as predictive
accuracy (\citeproc{ref-goodfellow2016deep}{Goodfellow, Bengio, and
Courville 2016}). Modern advances in computing have made it possible to
provide such models with ever-growing degrees of freedom to achieve that
task, which frequently allows them to outperform traditionally more
parsimonious models. Unfortunately, in doing so, models learn
increasingly complex and highly sensitive representations that humans
can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
(DL) revolution, Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2014}) showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AEs):
perturbed versions of data instances that yield vastly different model
predictions despite being ``imperceptible'' in that they are
semantically indifferent from their factual counterparts. Even though
some partially effective mitigation strategies have been proposed---most
notably \textbf{adversarial training}
(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
2015})---truly robust deep learning remains unattainable even for models
that are considered ``shallow'' by today's standards
(\citeproc{ref-kolter2023keynote}{Kolter 2023}).

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2014}) who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) who consider
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage CEs during
training and get adversarial examples essentially for free.

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
2023}). In this context, domain constraints can be imposed by simply
projecting counterfactuals back to the specified domain, if the previous
gradient step resulted in updated feature values that were
out-of-domain. Mutability constraints can similarly be enforced by
setting partial derivatives to zero to ensure that features are only
perturbed in the allowed direction, if at all.

@@

Our key outcome of interest is how well do models perform with respect
to explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
distance-based metric used by Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}),

@@

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
squared population maximum mean discrepancy
(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).

@@

\subsubsection{Plausibility}\label{sec-plaus}

Table~\ref{tbl-main} presents our main empirical findings. The top five
rows show the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. The following row shows the reduction
in implausibility as measured by Equation~\ref{eq-impl-div} and
aggregated across all test specifications of \emph{ECCo}. The following
two rows show the test accuracies for the model trained with CT and
conventionally trained models (``vanilla'').

@@

\caption{\label{tbl-main}Key explainability and predictive performance
metrics for all datasets. \textbf{Plausibility}: the top five rows show
the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. The following row shows the reduction
in implausibility as measured by Equation~\ref{eq-impl-div} and
aggregated across all test specifications. \textbf{Accuracy}: The
following two rows show the test accuracies for the models trained with
CT and the baseline. \textbf{Actionability}: The final row present the
average reduction in costs when imposing mutability constraints.}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
den Hengel. 2020. {``{Counterfactual Vision and Language Learning}.''}
In \emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
Recognition (CVPR)}, 10041--51.
\url{https://doi.org/10.1109/CVPR42600.2020.01006}.

\bibitem[\citeproctext]{ref-altmeyer2023explaining}
Altmeyer, Patrick, Arie van Deursen, and Cynthia C. S. Liem. 2023.
{``{Explaining Black-Box Models through Counterfactuals}.''} In
\emph{Proceedings of the JuliaCon Conferences}, 1:130. 1.

\bibitem[\citeproctext]{ref-altmeyer2024faithful}
Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
S. Liem. 2024. {``{Faithful Model Explanations through
Energy-Constrained Conformal Counterfactuals}.''} In \emph{Proceedings
of the Thirty-Eighth AAAI Conference on Artificial Intelligence},
38:10829--37. 10. \url{https://doi.org/10.1609/aaai.v38i10.28956}.

\bibitem[\citeproctext]{ref-augustin2020adversarial}
Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
{``{Adversarial Robustness on In- and Out-Distribution Improves
Explainability}.''} In \emph{Computer Vision -- ECCV 2020}, edited by
Andrea Vedaldi, Horst Bischof, Thomas Brox, and Jan-Michael Frahm,
228--45. Cham: Springer.

\bibitem[\citeproctext]{ref-balashankar2023improving}
Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
Chi, Jilin Chen, and Alex Beutel. 2023. {``{Improving Classifier
Robustness through Active Generative Counterfactual Data
Augmentation}.''} In \emph{Findings of the Association for Computational
Linguistics: EMNLP 2023}, 127--39. ACL.
\url{https://doi.org/10.18653/v1/2023.findings-emnlp.10}.

\bibitem[\citeproctext]{ref-bell2024fairness}
Bell, Andrew, Joao Fonseca, Carlo Abrate, Francesco Bonchi, and Julia
Stoyanovich. 2024. {``{Fairness in Algorithmic Recourse Through the Lens
of Substantive Equality of Opportunity}.''}
\url{https://arxiv.org/abs/2401.16088}.

@@

\bibitem[\citeproctext]{ref-du2019implicit}
Du, Yilun, and Igor Mordatch. 2020. {``{Implicit Generation and
Generalization in Energy-Based Models}.''}
\url{https://arxiv.org/abs/1903.08689}.

\bibitem[\citeproctext]{ref-franceschi2017forward}
Franceschi, Luca, Michele Donini, Paolo Frasconi, and Massimiliano
Pontil. 2017. {``{Forward and Reverse Gradient-Based Hyperparameter
Optimization}.''} In \emph{Proceedings of the 34th International
Conference on Machine Learning}, 1165--73. ICML'17. Sydney, NSW,
Australia: JMLR.org.

\bibitem[\citeproctext]{ref-frankle2018lottery}
Frankle, Jonathan, and Michael Carbin. 2019. {``{The Lottery Ticket
Hypothesis: Finding Sparse, Trainable Neural Networks}.''} In
\emph{International Conference on Learning Representations}.

\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
Freiesleben, Timo. 2022. {``{The Intriguing Relation Between
Counterfactual Explanations and Adversarial Examples}.''} \emph{Minds
and Machines} 32 (1): 77--109.

@@

\bibitem[\citeproctext]{ref-goodfellow2014explaining}
Goodfellow, Ian, Jonathon Shlens, and Christian Szegedy. 2015.
{``{Explaining and Harnessing Adversarial Examples}.''}
\url{https://arxiv.org/abs/1412.6572}.

@@

\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
Guidotti, Riccardo. 2022. {``{Counterfactual Explanations and How to
Find Them: Literature Review and Benchmarking}.''} \emph{Data Mining and
Knowledge Discovery} 38 (5): 2770--2824.
\url{https://doi.org/10.1007/s10618-022-00831-6}.

\bibitem[\citeproctext]{ref-guo2023counternet}
Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``{CounterNet:
End-to-End Training of Prediction Aware Counterfactual Explanations}.''}
In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
Discovery and Data Mining}, 577-\/-589. KDD '23. New York, NY, USA:
Association for Computing Machinery.
\url{https://doi.org/10.1145/3580305.3599290}.

@@

\bibitem[\citeproctext]{ref-joshi2019realistic}
Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
Joydeep Ghosh. 2019. {``{Towards Realistic Individual Recourse and
Actionable Explanations in Black-Box Decision Making Systems}.''}
\url{https://arxiv.org/abs/1907.09615}.

\bibitem[\citeproctext]{ref-kolter2023keynote}
Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)}. Los
Alamitos, CA, USA: IEEE Computer Society.
\url{https://doi.org/10.1109/SaTML54575.2023.00009}.

\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2017.
{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
Ensembles.''} In \emph{Proceedings of the 31st International Conference
on Neural Information Processing Systems}, 6405--16. NIPS'17. Red Hook,
NY, USA: Curran Associates Inc.

@@

\bibitem[\citeproctext]{ref-luu2023counterfactual}
Luu, Hoai Linh, and Naoya Inoue. 2023. {``{Counterfactual Adversarial
Training for Improving Robustness of Pre-trained Language Models}.''} In
\emph{Proceedings of the 37th Pacific Asia Conference on Language,
Information and Computation}, 881--88. ACL.
\url{https://aclanthology.org/2023.paclic-1.88/}.

@@

\bibitem[\citeproctext]{ref-ross2021learning}
Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024.
{``{Learning Models for Actionable Recourse}.''} In \emph{Proceedings of
the 35th International Conference on Neural Information Processing
Systems}. NIPS '21. Red Hook, NY, USA: Curran Associates Inc.

\bibitem[\citeproctext]{ref-sauer2021counterfactual}
Sauer, Axel, and Andreas Geiger. 2021. {``{Counterfactual Generative
Networks}.''} \url{https://arxiv.org/abs/2101.06046}.
